---
title: "Practical Machine Learning - Week 3 Course Project"
output: html_document
---

### Background and Objective

#####6 participants in a fitness project are asked to perform barbell lifts correctly and incorrectly in 5 different ways.The data is collected from accelerometers on the belt, forearm, arm, and dumbell of the participants.

#####The goal is to predict the manner in which they did the exercise. 

### Pulling the Data

##### Download the training and testing data from the given data sources links

```{r}
download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv",destfile="./acctrain.csv")

download.file("http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv",destfile="./acctest.csv")
```

##### Read the training and testing data from Working Directory. 
```{r}
acctrain<-read.csv("acctrain.csv",na.strings=c("NA","#DIV/0!",""))

acctest<-read.csv("acctest.csv",na.strings=c("NA","#DIV/0!",""))

```

### Explore and Clean Data


##### Training dataset has 19622 records and 160 columns and test dataset has 20 records and 160 columns.

```{r}

colNA<-colMeans(is.na(acctrain))

```
##### Lots of columns in the traing data have NA (missings) percentage greater than 97%. Imputing for missing values in these cases are futile so should remove these columns.

```{r}

acctrain2<-acctrain[,colMeans(is.na(acctrain))<.97]

```
 
##### After removing columns with high no. of missing values, Training dataset has 19622 records and 60 columns. 
 
##### Same operation on test data, Testing dataset has 20 records and 60 columns. 
 
```{r}

acctest2<-acctest[,colMeans(is.na(acctest))<.97]

```
 
####Keeping only the accelerometer related data in the training and test datasets. According to problem statement - 'your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants'

#####Note that dropping the 'Problem_ID' column from testing dataset as it is not required.


```{r}

acctrain3<-acctrain2[,c(grep("belt|forearm|arm|dumbbell",names(acctrain2)),ncol(acctrain2))]

acctest3<-acctest2[,c(grep("belt|forearm|arm|dumbbell",names(acctest2)))]
```


### Creating Training Data Partition to create a Cross-Validation Dataset using Caret library

##### Creating a 60%-40% split

```{r}
library(caret)

set.seed(15390)

inTrain <- createDataPartition(acctrain3$classe, p = .6, list=FALSE)
trainingdata <- acctrain3[inTrain,]
testingdata <- acctrain3[-inTrain,]

```

### Training the Model on 60% data

##### Run the Random Forest model using the reduced set of predictors to predict the 'classe' variable and evaluate how well it works predicting the training sample

```{r}
library(randomForest)

trctrl=trainControl(method="cv",10,repeats=5)

rfFit <- randomForest(classe ~., 
                       trControl=trctrl,
                       data=trainingdata, 
                       ntree=500, 
                       imp=TRUE)

varImpPlot(rfFit)

trpred <- predict(rfFit, trainingdata)

confusionMatrix( trainingdata$classe,trpred)

```

#### Accuracy of the model on training data looks good, now time for validating it on the hold-out sample.

### Validate the Model on 40% data

##### Fit the hold-out validation sample with the training RF model

```{r}
testpred <- predict(rfFit, testingdata)

confusionMatrix( testingdata$classe, testpred)

```

#### Obtained accuracy of 99.38%, which means an error of 0.62% in the hold-out sample. The model looks good enough to use on the 20 test data points.

###Predict the classe for the 20 test sample 

```{r}

finaltestpred <- as.character(predict(rfFit, acctest3))


pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(finaltestpred)

```